\section{Results}
\label{sec:results}

\subsection{Completed 100M-token architecture path}

\begin{table}[t]
\caption{Completed nominal 100M-token results read from archived CSV/JSON measurements. Throughput is one logged point at the final finite evaluation step on one H100. The GQA row lacks a realized configuration and used peak LR $3\times10^{-4}$; recent WVR and W/R/V runs used $4\times10^{-4}$.}
\label{tab:wvr-100m}
\centering
\small
\begin{tabular}{lrr}
\toprule
Mechanism & Held-out loss $\downarrow$ & Train tok/s $\uparrow$ \\
\midrule
Dense GQA (QKV) & \textbf{4.820476} & 111,355 \\
Shared additive associative context & 4.921405 & 124,363 \\
Stable Delta WVR & 4.898880 & 121,971 \\
Multi-timescale Delta WVR & 4.905554 & 120,496 \\
Collision-normalized WVR & \underline{4.895427} & 122,282 \\
Collision WVR + 1 lexical W/R/V layer & 4.892026 & 123,145 \\
Lexical-value WVR & 4.963167 & 123,113 \\
Lexical Forge V3 WVR & 4.900726 & 121,697 \\
Forge V3 + collision normalization & 4.897241 & 121,312 \\
Compact context + position address & 4.899072 & 122,764 \\
Strongly lexical W/R/V, 8/10 layers & 4.973798 & 118,546 \\
Contextual W/R/V, 8/10 layers & 4.852761 & 118,936 \\
Contextual W/R/V + RMSNorm, 10/10 & \textbf{4.706118} & 118,416 \\
\bottomrule
\end{tabular}
\end{table}

Collision-normalized WVR is the best completed fixed-state associative variant, improving Stable Delta by 0.003453 loss. Adding one W/R/V attention layer changes loss by 0.003401. The strongly lexical eight-layer condition reaches 4.973798, while the contextual eight-layer hybrid reaches 4.852761. The final jointly changed condition removes the two remaining WVR layers and adds per-head RMS normalization, reaching 4.706118 at 118,416 tokens/s. These path comparisons do not isolate lexical residuals, normalization, or layer replacement as causes.

The final row is 0.114358 lower in loss, and its single logged throughput point is 6.34\% higher than the archived GQA point. This is not an architectural superiority or reproducible speedup estimate: peak learning rate differs and the GQA realized configuration is missing. A fully archived GQA rerun at $4\times10^{-4}$ is required before declaring a controlled winner.

The lexical forge reuses the rank-16 lexical-object table through a $16\rightarrow4\rightarrow128$ network. Its strong short-proxy improvement did not transfer: Forge V3 reached 4.900726 at full budget, and combining it with collision normalization reached 4.897241 rather than improving the collision-only result. A lexical payload was substantially worse at 4.963167, supporting the design decision to retain $\mathbf{v}_t=\mW_v\vh_t$. Independent and weakly coupled read/write forge variants were rejected in paired proxies; their raw JSON artifacts are retained as negative ablations rather than promoted to the canonical implementation.

\subsection{Proxy reliability}

Short matched proxies were useful for detecting catastrophic failures but were not reliable rankers of final 100M-token loss. Multi-timescale Delta, Lexical Forge V3, and Forge V3 plus collision normalization each looked favorable in a short proxy and failed to improve the corresponding full-budget control. Consequently, no architectural claim in Table~\ref{tab:wvr-100m} is based on a proxy. The compact context-plus-position occurrence address also failed to improve collision normalization at full budget (4.899072 versus 4.895427).

\subsection{Longer-budget convolutional run}

The separate historical 1B-token attention-free convolutional run contains 98,194,244 parameters and reached held-out loss 4.180625 at step 30,000 with measured throughput 113,621 tokens/s. Because its token budget differs by a factor of ten, it is not used to rank the 100M-token QKV and WVR mechanisms.

\subsection{Incremental equivalence and fixed state}

FP32 unit tests reproduce full-sequence outputs under token-by-token incremental execution at tolerance $10^{-5}$ for the convolution and WVR primitives. The archived BF16 checkpoint profiles show larger numerical differences (approximately 0.02--0.17 maximum absolute error), so exact checkpoint-level BF16 equivalence is not claimed. For WVR, the recurrent state is one rank-128 association matrix, the previous write address, and---for collision normalization---one rank-128 diagonal load vector. The occurrence-address ablation adds one scalar absolute-position counter. These state shapes do not grow with generated context length.

We audit the realized exported convolutional checkpoint rather than inferring the absence of attention from configuration alone. Across 182 tensors, the name-and-shape audit finds no query, key, value, or fused-QKV projection. Compatibility paths such as \texttt{self\_attn} do not correspond to an attention operation. This structural audit is not treated as a substitute for runtime tests.

The archived associative-recall and induction probes report zero accuracy for the evaluated attention-free checkpoints at every tested distance. The available GQA synthetic artifact does not document an identical protocol, so we do not report a quantitative positive-control comparison. The negative attention-free diagnostic alone does not establish instance-selective retrieval.
