\section{Limitations and Broader Impact}

The current study is limited to approximately 100M parameters, one tokenizer, one web-text corpus, and one seed for the principal runs. A favorable result does not establish scaling behavior or universal superiority to GQA. Hardware throughput is implementation- and device-specific and must not be generalized from one H100.

The current study does not establish production inference throughput. W/R/V retains W/V tensors whose size grows with context, like a grouped KV cache. Long-context prefill, time to first token, serving latency distributions, and CUDA Graph replay remain outside the present claims.

Token-identity conditioning may encode corpus-specific frequency and lexical biases. FineWeb-Edu inherits limitations of large web corpora, including uncertain provenance, representational imbalance, and potentially harmful content. The proposed architecture does not remove these risks. Efficiency gains can also lower the cost of deploying low-quality or harmful models. We report data provenance, held-out construction, and deployment constraints so these tradeoffs can be assessed.

Fixed-state WVR does not retain individually addressable past token occurrences. Repeated lexical addresses and projected-context collisions are compressed into one recurrent matrix. Collision normalization mitigates coordinate saturation but does not recover overwritten instances. These negative controls motivate contextual W/R/V attention; they are not properties of the final model.

The strongest comparison is not optimizer-matched: GQA used peak LR $3\times10^{-4}$ and final W/R/V used $4\times10^{-4}$. The final run also saved no checkpoint, preventing post-training inspection of lexical write gates. Consequently, the current evidence cannot isolate gains from learning rate, per-head normalization, implementation details, or the lexical residual. Required controls are GQA at $4\times10^{-4}$ and W/R/V with the lexical residual fixed off, both with saved checkpoints and preferably multiple seeds. Finally, short proxies were not reliable predictors of full-budget ordering; they should remain failure screens rather than winner selectors.
